---
title: "Week 1 lab: Introduction to Tidyverse"
author: Hainan Xu
format: pdf
editor: visual
---

### Lab Exercises

**1. Plot the ratio of female to male mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme**

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
```

```{r message=FALSE,warning=FALSE}
#import data
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
```
```{r}
dm1<-dm |> mutate(fm_ratio=Female/Male) |> 
  filter(Age %in% c(10,20,30,40)) |> 
  select(Year,Age,fm_ratio) 
dm1 |> ggplot(aes(x=Year, y=fm_ratio, color=Age)) +
  geom_line() +
  labs(title="The Ratio of Female to Male Mortality Rates Over Time ",y="Ratio of Female to Male Mortality Rates" )+
  theme_bw()
```

**2. Find the age that has the lowest female mortality rate each year**

The dataframe below contains the ages that has the lowest female mortality rate each year. In some years, there are several ages that has the lowest female mortality rate.

```{r}
dm2 <- dm |> select(-Male,-Total) |>  
  group_by(Year) |>  
  filter(Female==min(Female,na.rm = TRUE))
head(dm2)
```

**3. Use the \`summarize(across())\` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.**

The output below show the first 6 rows of the standard deviation of mortality rates by age for male, female and total populations.

```{r}
dm3 <- dm |> group_by(Age)|> 
  summarize(across(Female:Total,~sd(.x,na.rm = TRUE)))
head(dm3)
```

**4. The Canadian HMD also provides population sizes over time. Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: \`left_join\` will probably be useful here.**

Comment: The observed output indicates a general decreasing trend in the average mortality rates for both women and men over time, with a slight rebound since 2007; furthermore, males exhibit a consistently higher mortality rate compared to females.

```{r}
#import the additional data
hmd=read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt",skip = 2, col_types = "dcddd")
#combine the data
dm4 <- dm |> left_join(hmd, by = c("Year","Age"))
dm4.1<- dm4 |> mutate(female_mortality=Female.x*Female.y,
              male_mortality=Male.x*Male.y) |>
  group_by(Year) |>
  summarise(#denominator of weighted average 
            female_death_total=sum(female_mortality,na.rm = TRUE),
            male_death_total=sum(male_mortality, na.rm = TRUE),
            #numerator of weighted average
            female_total=sum(Female.y),
            male_total=sum(Male.y)
            ) |>
  mutate(Female_Average=female_death_total/female_total,
         Male_Average=male_death_total/male_total) |>
  select(Year,Female_Average, Male_Average)
#plot
dm4.1 <- dm4.1 |> pivot_longer(cols=c("Female_Average","Male_Average"), names_to = "Sex", values_to="Average_Mortality_Rate")
ggplot(dm4.1)+
  geom_line(aes(x=Year,y=Average_Mortality_Rate, color= Sex))+
  scale_color_manual(
    values = c("Female_Average" = "pink", "Male_Average" = "blue"),
    labels = c("Female", "Male")) +
  labs(title = "The Average Motality Rate from 1921 - 2019", y="Average Mortality Rate")+
    theme_bw()
```

**5. Write down using appropriate notation, and run a simple linear regression with logged mortality rates as the outcome and age (as a continuous variable) as the covariate, using data for females aged less than 106 for the year 2000. Interpret the coefficient on age.**

We fit the model

$$
log(y_i)= \beta_0 +\beta_1 x_i+\epsilon_i
$$

where $y_i$ is the mortality rate of the whole population in Ontario in 2000, $x_i$ is the age, $\epsilon_i\sim (0,\sigma^2)$. According to the model summary shown below, the coefficient on age $\beta_1$ is 0.084, which means on average, the mortality rate increases $8.4\%$ for one year increase in age.

```{r}
dm$Age=as.numeric(dm$Age)
dm5<- dm |> filter(Year ==2000, Age < 106) |> na.omit()
model<-lm(log(Total) ~ Age, data= dm5)
summary(model)
```
